---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(datasetjson)
```

# Using **datasetjson**

**datasetjson** works by allowing you to take a data frame and apply the necessary attributes required for the CDISC Dataset JSON. The goal is to make this experience simple. Before you can write a Dataset JSON file to disk, you first need to build the Dataset JSON object. An example call looks like this:

```{r sample_call, eval=FALSE}
ds_json <- dataset_json(iris, "IG.IRIS", "IRIS", "Iris", iris_items)
```

This is the minimum information required to provide to create a `datasetjson` object. 

The parameters here can be described as follows:

- The input data frame `iris`
- The `item_id`, which can be described as the "Object of Dataset", which is a key value is a unique identifier for the dataset, corresponding to ItemGroupDef/@OID in Define-XML.
- `name`, which is the dataset name
- `label`, which is the dataset label, and finally
- `items`, which is the variable level metadata for your dataset.

The `items` parameter is special here, in that you provide a data frame with the necessary variable metadata. Take a look at the `iris_items` data frame. 

```{r iris_items}
iris_items
```

This data frame has 7 columns, 4 of which are strictly required:

| **Variable Name** | **Description**                                                                                                 |
|-------------------|-----------------------------------------------------------------------------------------------------------------|
| OID               | Unique identifier for Variable. Must correspond to ItemDef/@OID in Define-XML.                                  |
| name              | Name for Variable                                                                                               |
| label             | Label for Variable                                                                                              |
| type              | Data type for Variable                                                                                          |
| length            | Length for Variable                                                                                             |
| keySequence       | Indicates that this item is a key variable in the dataset structure. It also provides an ordering for the keys. |
| displayFormat     | Display format supports data visualization of numeric float and date values.                                    |

The data within this dataframe ultimate populates the `items` element of the Dataset JSON file. The OID, name, label, and type columns are all required and must be populated for each variable. Note that the type column has a list of allowable values:

- string
- integer
- float
- double
- decimal
- boolean

This information must be provided directly by the user. Note that no type conversions of your data are performed by the `datasetjson` package. The displayFormat column inherently refers to display formats used within SAS. 

## Setting Other Data Attributes
