---
title: "Numeric Precision"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{precision}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Numeric precision and issues with floating point decimals are a common problem when working with data. Dataset JSON is not immune to these issues. Instead of writing out direct binary representations of floating point numbers, which vary depending on the system being used and the standard followed, Dataset JSON writes out character representations of these numbers. Whenever a number is serialized to text and parsed back, precision can be lost.

**{datasetjson}** parses and serializes numbers in C, using the [yyjson](https://github.com/ibireme/yyjson) library that ships with the package. Numbers are written using the shortest text that reads back as the same value, and read straight into a double without passing through R's `as.double()`. The practical effect is that a value survives a write and read unchanged:

```{r example}
library(datasetjson)
library(dplyr)

test_df <- head(iris, 5)
test_df['float_col'] <- c(
  143.66666666666699825,
  2/3,
  1/3,
  165/37,
  6/7
)

test_items <- iris_items |> bind_rows(
  data.frame(
    itemOID = "IT.IR.float_col",
    name = "float_col",
    label = "Test column long decimal",
    dataType = "float"
  )
)

dsjson <- dataset_json(
  test_df, 
  item_oid = "test_df",
  name = "test_df",
  dataset_label = "test_df",
  columns = test_items
)

json_out <-write_dataset_json(dsjson)

out <- read_dataset_json(json_out)

test_df$float_col - out$float_col
```

Every difference is exactly zero: the values that came back are bit-for-bit the values that went in, and that holds across the full range of double precision, from `1e-300` to `1e+300`.

This has not always been true. Before version 0.4.0 the numbers were converted to text by the JSON library in use at the time, at a fixed six decimal places. Values were rounded at around the seventh decimal, and anything smaller than roughly `5e-7` was read back as exactly `0`. If you have files written by an earlier version of the package, the data in them is fine - the loss happened on read, so re-reading them with this version recovers the original values.

## The decimal data type

The Dataset JSON standard also offers a "decimal" type. From the [user guide](https://wiki.cdisc.org/display/PUB/Precision+and+Rounding):

> ## Decimal Data Type
>
> Although the pilot findings on precision and rounding did not point to a problem with Dataset-JSON, the Dataset-JSON Team opted to add the Decimal datatype. The Decimal datatype has been available in ODM for many years. The basic premise for this datatype is to represent the number in Dataset-JSON as a string (a quoted set of numeric characters) to prevent JSON libraries from interpreting the number as a float before the software application gets access to it.
>
> To use decimal in Dataset-JSON, set the dataType to decimal and the targetDataType to decimal. This instructs conversion software to convert the number it reads from a native dataset into a string in Dataset-JSON. It also instructs the receiver to convert the number as a string into the decimal datatype or closest approximation available in the receiving technology. Note that not all technologies support an explicit decimal datatype.

`write_dataset_json()` supports this through `float_as_decimals`. When set, float columns are written as `decimal`/`decimal` and the values are quoted, and `read_dataset_json()` converts them back to numeric on read per the specification.

```{r conversion}
json_out <- write_dataset_json(dsjson, float_as_decimals = TRUE)

out <- read_dataset_json(json_out)

test_df$float_col - out$float_col
```

The result is the same, because both paths are exact. **`float_as_decimals` is an interoperability choice, not a precision workaround.** Reach for it when the system receiving your file needs numbers presented as strings; leave it off otherwise. It remains off by default for two reasons:

- Writing the `decimal` type is an extra step the consuming system has to be aware of, and Dataset JSON is still a young standard
- Quoted numbers make for larger files with no gain in fidelity

## The `digits` argument

`digits` controls how many significant digits are used when rendering decimal strings. It defaults to `NULL`, meaning the shortest representation that still reads back as the same value - which is what makes the decimal path lossless, and keeps the output tidy:

```{r digits}
cat(write_dataset_json(dsjson, float_as_decimals = TRUE, pretty = TRUE))
```

Supplying a number fixes the precision instead. That is occasionally useful for matching another system's output, but it is lossy whenever the value needs more digits than you allow:

```{r digits_fixed}
fixed <- read_dataset_json(
  write_dataset_json(dsjson, float_as_decimals = TRUE, digits = 8)
)

test_df$float_col - fixed$float_col
```

Note that 17 significant digits, not 16, is the round-trip guarantee for a double. If you do set `digits`, anything below 17 risks losing the last bits of a value.
